In a random sample of 100 patients at a clinic, you would like to test whether the mean RDI is x or more using
a one sided 5% type 1 error rate. The sample mean RDI had a mean of 12 and a standard deviation of 4.
What is the largest value of x (H0:μ=x versus Ha:μ>x) would you reject for?
Give at least one decimal place in your answer.
```{r}
mu =12
s = 4
n =100
mu+qt(0.95, n-1)*s/sqrt(n) 
```
A pharmaceutical company is interested in testing a potential blood pressure lowering medication. Their first examination considers only subjects that received the medication at baseline then two weeks later. The data are as follows (SBP in mmHg)
Subject    Baseline	Week 2
1	140	132
2	138	135
3	150	151
4	148	146
5	135	130
Consider testing the hypothesis that there was a mean reduction in blood pressure? Give the P-value for the associated two sided test. Give at least 3 decimal places in your answer and express your P-value as a proportion and not a percentage.
```{r}
x1 = c(140, 138, 150, 148, 135)
x2 = c(132, 135, 151, 146, 130)
t.test(x2-x1)
```

A sample of 9 men yielded a sample average brain volume of 1,100cc and a standard deviation of 30cc. What is the complete set of values of μ0 that a test of H0:μ=μ0 would fail to reject the null hypothesis in a two sided 5% Students t-test?

```{r}
mu =1100
sd = 30
n=9
mu+c(-1,1)*qt(0.975, n-1)*sd/sqrt(n) 
```
Researchers conducted a blind taste test of Coke versus Pepsi. Each of four people was asked which of two blinded drinks given in random order that they preferred. The data was such that 3 of the 4 people chose Coke. Assuming that this sample is representative, report a P-value for a test of the hypothesis that Coke is preferred to Pepsi using a one sided exact test. Give the answer as either a proportion to two decimal places or a percentage to the nearest percentage point.
