Title:  Explore the NOAA storm Database, and answer what types of weather events are harmful?
=========================================


Synopis
---------------------------------
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in *fatalities*, *injuries*, and *property damage*, and preventing such outcomes to the extent possible is a key concern.

This project  explores the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, which tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
Exploring this database, we can answer two the following questions:
* Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?

Data Processing
--------------------------------------

The data for this project come be download from [web site](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). It is in the form of a comma-seperated-value file, compressed via the bzip2 algorithm.

### Read data file 
```{r readDatafile,echo=TRUE,cache=TRUE}

fname="StormData.csv"
if(!file.exists(fname)){
    dataURL="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    temp <-"StormData.csv.bz2"
    download.file(dataURL, temp, method="curl")
    library(R.utils)
    bunzip2(temp,"StormData.csv",remove=FALSE)
    unlink(temp)
}
data <- read.csv(fname)
```


First, brief look at data and find out what kinds of information are available.
```{r data_summary, echo=TRUE}
dim(data)
#summary(data)
head(data)
str(data)
ncolumns <- ncol(data)
```


#### Extract select columns
Although the dataset has **`r ncolumns`** fields, only a few are intereting for this project.
```{r selectField,echo = TRUE}
selectFields <- c("STATE","EVTYPE","BGN_DATE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")
data <- data[,selectFields]
```

#### Restrict to US states
Here we are only interesting data on US, so using R's built-in `state.abb` dataset to restrict the records of US *STATE*s.
```{r restrictUS,echo=TRUE}
data <- data[data$STATE %in% state.abb,]
```

#### Covert *BGN_DATE* field into a new beginDate date class.
```{r convertdates, cache=TRUE}
#data$BGN_DATE <- as.Date(as.POSIXlt(data$BGN_DATE,format="%m/%d/%Y %H:%M:%S"))
data$beginDate <- as.Date(as.POSIXct(data$BGN_DATE,format="%m/%d/%Y %H:%M:%S"))
```

#### Add new property damage variable **propertyDamage**
*'propertyDamage'*, is defined as the product of *PROPDMG* and *PROPDMGEXP*,
where "H" for hundres, “K” for thousands, “M” for millions, and “B” for billions. 
```{r propertyDamage,echo=TRUE}
data$PROPDMGEXP <- toupper(data$PROPDMGEXP)
data$propertyDamage <- ifelse(data$PROPDMGEXP == "B", data$PROPDMG*1E9,
                              ifelse(data$PROPDMGEXP == "M", data$PROPDMG*1E6,
                                     ifelse(data$PROPDMGEXP == "K", data$PROPDMG*1E3,
                                         ifelse(data$PROPDMGEXP == "H", data$PROPDMG*1E2,
                                                data$PROPDMG
                                             ))))

summary(data$propertyDamage)
```

#### Another new variable, *`cropDamage`*, defined similar ass 'propDamage'.
```{r}
data$CROPDMGEXP <- toupper(data$CROPDMGEXP)
data$cropDamage <- ifelse(data$CROPDMGEXP == "B", data$CROPDMG*1E9,
                              ifelse(data$CROPDMGEXP == "M", data$CROPDMG*1E6,
                                     ifelse(data$CROPDMGEXP == "K", data$CROPDMG*1E3,
                                         ifelse(data$CROPDMGEXP == "H", data$CROPDMG*1E2,
                                                data$CROPDMG
                                                ))))                                                
summary(data$cropDamage)
```

#### Clean *EVTYPE* variable
```{r cleanEvtype, echo=TRUE}
nEVTYPE <- length(unique(data$EVTYPE))
```
There are **`r nEVTYPE`** types of events recorded in this database, listed as above. The unique number is too large to manage without grouping. Here we refer the group categories found in the [2009 Annual Summaries](http://www.ncdc.noaa.gov/oa/climate/sd/annsum2009.pdf) on page 3. There are 7 categories: Convection, Extreme Temperature, Flood, Marine, Tropical Cyclones, Winter and Other. In each category has many sub-types. For the sake of simplicity, marine and tropical cyclones are put into Other.

For the event types fall into the category of **Convection**.
```{r evtConvection,echo=TRUE}
Lightning    <- "\\bL\\S+?G\\b"
Tornado      <- "(NADO)|(\\bTOR\\S+?O\\b|(\\bFUN))"
Thunderstorm <- "THUNDERSTORM|TSTM"
Wind         <- "(WIND)|(WND)"
Hail         <- "HAIL"
regex <- paste(Lightning, Tornado, Thunderstorm, Wind, Hail, sep="|")
data$eventConvection <- grepl(regex, data$EVTYPE, ignore.case = TRUE)
```


For variations of **Cold** and **Heat**, list the event types that fall into the category of **Extreme Temperatures**.
```{r evtExtremTemp,echo=TRUE}
regex <- "COLD|HEAT"
data$eventExtremeTemp <- grepl(regex, data$EVTYPE, ignore.case = TRUE)
```

For variations of **Flood** and **Rain**,list the event types that fall into the category of **Flood**.
```{r evtFlood,echo=TRUE}
Flood <- "(\\bFL\\S+?D)"
Rain  <- "RAIN|PRECIP|SHOWER"
regex <- paste(Flood, Rain, sep="|")

data$eventFlood <- grepl(regex, data$EVTYPE, ignore.case = TRUE)
```

For variations of **Snow**, **Ice**, **Freeze**, or **Winter Weather**,
list the event types that fall into the category of **Winter**.
```{r evtWinter,echo=TRUE}
regex <- "(SNOW)|(ICE)|(ICY)|(FREEZ)|(WINT)"
data$eventWinter <- grepl(regex, data$EVTYPE, ignore.case = TRUE)
```


For event types that don't satisfy any one of the aboves,
create an **Other** indicator for ungrouped event types.
```{r evtOther,echo=TRUE}
where <- expression(data$eventConvection == FALSE & data$eventExtremeTemp == FALSE & data$eventFlood == FALSE & data$eventWinter == FALSE)
data$eventOther <- eval(where)
```

#### Categorize event types
Now we can set up a categorization hierarchy of event types.
Notices 'EVTYPE' records can have multiple events, E.g., *WINTER STORM/HIGH WINDS* which set both *eventConvection* and *eventWinter*.
A crosstabulation for the event type categories is below.
```{r crossTable,echo=TRUE}
d1<-aggregate(propertyDamage~eventConvection+eventExtremeTemp+eventFlood+eventWinter+eventOther, data=data, sum)
d1
```
To simplify the problem,  we choose higher categories outrank lower categories, 
The hierarchy is as follows.
1. Convection (including lightning, tornado, thunderstorm, wind, and hail)
2. Extreme temperature (including hot and cold)
3. Flood (including flood, flash flood, rain)
4. Winter (including snow, ice, freeze, or winter weather)
5. Other
That is, the example event type of *WINTER STORM/HIGH WIND* would only be considered as *Convection* category.
```{r evtCategory,echo=TRUE}
data$eventCategory <- ifelse(data$eventConvection, "Convection",
                             ifelse(data$eventExtremeTemp, "ExtremeTemp",
                                    ifelse(data$eventFlood, "Flood",
                                           ifelse(data$eventWinter, "Winter",
                                                  ifelse(data$eventOther, "Others",
                                                         NA
                                                         )))))


table(data$eventCategory)
```



#### Restrict date range

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete. The date ranges for each category are listed below.
Filter the data to include records including all considered event categories.
```{r restrictDateRange,echo=TRUE}
d1<- aggregate(beginDate~eventCategory, data, min)
d2<- aggregate(beginDate~eventCategory, data, max)

cbind(d1,end_BeginDate=d2$beginDate)
#boxplot(log(propertyDamage)~eventCategory, data=data[data$propertyDamage>0,])
library(lubridate)
minYear <- max(year(d1$beginDate))
maxYear <- max(year(d2$beginDate))
data <- data[year(data$beginDate)>=minYear,]
```


### Results
* Across the United States, which types of events are most harmful with respect to population health?
Here we consider **FATLITIES** and **INJURIES**.

```{r resultsFAT,echo=TRUE}
labels <- c("Convection", "Extreme temperature", "Flood", "Winter", "Other")
d_Fatalities <- data[data$FATALITIES>0,]
tab_Fatalities<-aggregate(FATALITIES~eventCategory,d_Fatalities,sum)
maxFcategory    <- max(tab_Fatalities$FATALITIES)
maxFEvtCategory <- labels[which(tab_Fatalities$FATALITIES==maxFcategory)]
library(ggplot2)
ggplot(data=tab_Fatalities, aes(x=eventCategory, y=FATALITIES,fill=FATALITIES))+ geom_bar(stat="identity")
```
The most harmful event category is **`r maxFEvtCategory`** with the total dealth of **`r maxFcategory`** between *`r minYear`* and *`r maxYear`*. 

Here list the top 5 harmful events with largest fatalities in US.
```{r top5Fat,echo=TRUE}
data[order(data$FATALITIES,decreasing=T),][1:5,c("STATE","EVTYPE","beginDate","eventCategory","FATALITIES")]
```

```{r inj,echo=TRUE}
d_Injuries   <- data[data$INJURIES>0,]
tab_Injuries<-aggregate(INJURIES~eventCategory,d_Injuries,sum)
maxFcategory    <- max(tab_Injuries$INJURIES)
maxFEvtCategory <- labels[which(tab_Injuries$INJURIES==maxFcategory)]
ggplot(data=tab_Injuries, aes(x=eventCategory, y=INJURIES,fill=INJURIES))+ geom_bar(stat="identity")
```
The most harmful event category is **`r maxFEvtCategory`** with the total injuries of **`r maxFcategory`** between *`r minYear`* and *`r maxYear`*. 

Here list the top 5 harmful events with largest injuries in US.
```{r top5Injuries,echo=TRUE}
data[order(data$INJURIES,decreasing=T),][1:5,c("STATE","EVTYPE","beginDate","eventCategory","INJURIES")]
```

* Across the United States, which types of events have the greatest economic consequences?
```{r RpropertyDMG,echo=TRUE}
d_DMG <- data[data$propertyDamage>0,]
tab_DMG <- aggregate(propertyDamage~eventCategory,d_DMG,sum)
maxFcategory <- max(tab_DMG$propertyDamage)
maxFEvtCategory <- labels[which(tab_DMG$propertyDamage==maxFcategory)]
maxFcategory <- maxFcategory/1E9
ggplot(data=tab_DMG, aes(x=eventCategory, y=propertyDamage,fill=propertyDamage))+ geom_bar(stat="identity")
```
The most property damaged event category is **`r maxFEvtCategory`** with the total property damage of **$`r maxFcategory`B** between *`r minYear`* and *`r maxYear`*.

Here list the top 5 property damaged event in US.
```{r top5Damage,echo=TRUE}
#data[which(data$propertyDamage==max(data$propertyDamage)),c("STATE","EVTYPE","beginDate","eventCategory","propertyDamage")]
data[order(data$propertyDamage,decreasing=T),][1:5,c("STATE","EVTYPE","beginDate","eventCategory","propertyDamage")]
```


