
Peer Assessments -- Regression Models
==============================================================================

Looking at a data set of a collection of cars, mtcars, we are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome), especially the below two questions:
* “Is an automatic or manual transmission better for MPG?”
* “Quantifying how different is the MPG between automatic and manual transmissions?”

### Loading dataset
```{r loadData,echo=TRUE}
library(datasets); data(mtcars); require(stats); require(graphics)
?mtcars
```

### Explory data

correlation matrix
```{r cor,echo=TRUE}
corMat <- cor(mtcars)
library(corrplot)
#corrplot(corMat, method="circle",tl.cex=0.5)
corrplot.mixed(corMat)
```

It shows that the varialbes in mtcars dataset are strongly correlated.

### Look the relationship between *am* and other variables
```{r amplots,echo=FALSE,fig.width=10,fig.height=16}
mtcars<-transform(mtcars, am=factor(am))


require(datasets);data(mtcars)
require(stats); require(graphics)
par(mfrow=c(3,2))
boxplot(mpg ~ am, data = mtcars,
        xlab = "Transimission (0=automatic, 1= manual)", ylab = "MPG",
        main = "mtcars data", varwidth = TRUE, col = "lightgray")

boxplot(cyl ~ am, data = mtcars,
        xlab = "Transimission (0=automatic, 1= manual)", ylab = "Number of cylinders",
        main = "mtcars data", varwidth = TRUE, col = "lightgray")

boxplot(disp ~ am, data = mtcars,
        xlab = "Transimission (0=automatic, 1= manual)", ylab = "Displacement",
        main = "mtcars data", varwidth = TRUE, col = "lightgray")

boxplot(drat ~ am, data = mtcars,
        xlab = "Transimission (0=automatic, 1= manual)", ylab = "Rear axle ratio",
        main = "mtcars data", varwidth = TRUE, col = "lightgray")

boxplot(wt ~ am, data = mtcars,
        xlab = "Transimission (0=automatic, 1= manual)", ylab = "Weight (lb/1000)",
        main = "mtcars data", varwidth = TRUE, col = "lightgray")

boxplot(gear ~ am, data = mtcars,
        xlab = "Transimission (0=automatic, 1= manual)", ylab = "Number of forward gears",
        main = "mtcars data", varwidth = TRUE, col = "lightgray")


par(mfrow=c(1,1))
```
---

### Model strategory
* Basic linear regession 
$$ mpg_i = \beta_0 + am_{i1} \beta_1 + \epsilon_{i} $$
* Multivariate linear regession with all existing variables
$$ mpg_i = \sum_{k=1}^p X_{ik} \beta_j + \epsilon_{i} $$
* Two variates linear regession ()
$$ mpg_i = \beta_0 + am_{i1} \beta_1 + wt_{i1} \beta_2 + \epsilon_{i} $$


### linear model1
$$
mpg_i = \beta_0 + am_{i1} \beta_1 + \epsilon_{i}
$$
* $\beta0 = 17.1$ is the mpg of $am=0$ (automatic transimission)
* $\beta0 +\beta1 = 24.4$ is the mpg of $am=1$ (manual transimission)
* Quantifying how different is the MPG between automatic and manual transmissions?
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
manual transimission is 7.2 mpg better.


```{r model1,echo=TRUE}
 fit1 <- lm(mpg~am, data=mtcars)
# e<-resid(fit1)
 boxplot(mtcars$am,resid(fit1), xlab="Transimission", ylab="resid")

 summary(fit1)$coef
```
---


### linear model2
$$
mpg_i = \sum_{k=1}^p X_{ik} \beta_j + \epsilon_{i} \\
    = \beta_0 + cyl\beta_1 + disp\beta_2 +hp\beta_3 + drat\beta_4
    + wt\beta_5 +qsec\beta_6 +vs\beta_7 + am\beta_8 + gear\beta_9+carb \beta_10 + \epsilon_{i}
$$
* the coeffient of transimmison is $\beta8 =2.52\pm2.06$ with t-value $1.225$. When all other variables are equal, the manual transimission car is 2.52 mpg better (p-value 0.234).

```{r model2,echo=TRUE}
fit2 <- lm(mpg~., data=mtcars)
summary(fit2)$coef
```


### linear model3
$$ mpg_i = \beta_0 + am_{i1} \beta_1 + wt_{i1} \beta_2 + \epsilon_{i} $$
* $\beta_0 =37.2\pm 3.1 $, $\beta_1 = -0.02\pm 1.55$, $beta_3 = -5.35\pm 0.79$.
* The coefficent of am is not significant. no observable effect for automatic/manual transimission.
```{r}

fit3 <- lm(mpg~ am+wt, data=mtcars)
e <-summary(fit3)$coefficients
e

plot(mpg~wt,col=am+2, data=mtcars)
abline(c(e[1,1], e[2,1] ), col="red", lwd=3 )
abline(c(e[1,1], e[2,1]+e[3,1]), col="black", lwd=3)
```

## Summary
Two Questions
* “Is an automatic or manual transmission better for MPG?”
* “Quantifying how different is the MPG between automatic and manual transmissions?”

Answers are model -dependent
* Model1 and model2 show manual transimission are better, the quantity is different.
* Model3 show no significatn different 

```{r}
anova(fit1,fit3,fit2)
```
Anova table shows:
* Model3 is better than model1. p-value is less than 0.05, rejecting the null hypthosis, which means the *wt* variable may be necessary;
* Model3 is also better than model2. p-value is larger than 0.05, fail to reject the null hypthosis, means the other variables may not be necessary.