
Practicle Machine learning project -- predict the weavable activities
====================================

## introduction
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible
to collect a large amount of data about personal activity relatively inexpensively.
These type of devices are part of the quantified self movement – a group of
enthusiasts who take measurements about themselves regularly to improve their
health, to find patterns in their behavior, or because they are tech geeks.
One thing that people regularly do is quantify how much of a particular activity
they do, but they rarely quantify how well they do it.

In this project, the goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website
here: http://groupware.les.inf.puc-rio.br/har
(see the section on the Weight Lifting Exercise Dataset). 


### Data
==========================================================
The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. 

The data can be downloaded using the below R script.
```{r download_file,echo=TRUE}
downloadFiles<-function(
    dataURL="", destF="t.csv"
    ){
        if(!file.exists(destF)){
            download.file(dataURL, destF, method="curl")
        }else{
            message("data already downloaded.")
        }
    }
```

### loading training and testing dataset
```{r activity_data,echo=TRUE}
trainURL<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testURL <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
downloadFiles(trainURL, "pml-training.csv")
downloadFiles(testURL, "pml-test.csv")

training <- read.csv("pml-training.csv",na.strings=c("NA",""))
testing <-read.csv("pml-test.csv",na.strings=c("NA",""))
```

First look of training data
```{r}
dim(training)
str(training)
```

The outcome is "classe" variable
```{r}
table(training$classe)
```
The train dataset has 160 variables 19622 observations.

### remove missing values, is.na() or ""
```{r}
var <- names(training)[apply(training,2,function(x) table(is.na(x))[1]==19622)]   
train2<- training[,var]
test2 <- testing[,var[-length(var)]]
```

# discards unuseful predictors, only considering numeric variable from HAR sensor
```{r}
removeIndex <- grep("timestamp|X|user_name|new_window|num_window",names(train2))
train3 <- train2[,-c(removeIndex, length(train2))]
test3  <- train2[,-c(removeIndex, length(train2))]
```


Check the near Zero covariates and correlation matrix
removing zero covariates
```{r}
library(caret)
nzv <- nearZeroVar(train3, saveMetrics=TRUE)
nzv
nzv[nzv$nzv,]
```

highly correlated covariates.
```{r}
corrM <- cor(train3)
library(corrplot)
corrplot(corrM, method="circle",tl.cex=0.5)
```

remove highly correlated
```{r}
highCorr <- findCorrelation(corrM, cutoff = .75)     # high correlation
predictor <- train3[, -highCorr]            # dataframe of train predictors
filtered.test4 <- test3[, -highCorr]        # dataframe of test predictors
classe<-train2$classe                                       # target variable
trainData<-cbind(classe,predictor)    
```

Random Forest, to predic
```{r}
library(randomForest)
rfModel <- randomForest(train2$classe ~ .,data = trainData,importance = TRUE,ntrees = 500)
print(rfModel)

par(mar=c(3,4,4,4))                               
plot(rfModel)  
varImpPlot(rfModel,cex=.5)  
out.test<-predict(rfModel,filtered.test4)        #
```

```{r}
# make model
tc <- trainControl("cv", 10, savePred=T)
modFit <- train(train2$classe ~.,data = trainData,method="rf", trainControl=tc)
modFit
modFit$finalModel
out.test2<-predict(modFit,filtered.test4)   
out.test2[1:20]
```

### saving the output
```{r}
#answers<- as.vector(out.test)
answers = rep("A", 20)
pml_write_files = function(x){
#  n = length(x)
    n=20
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(answers)
```


