Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible
to collect a large amount of data about personal activity relatively inexpensively.
These type of devices are part of the quantified self movement – a group of
enthusiasts who take measurements about themselves regularly to improve their
health, to find patterns in their behavior, or because they are tech geeks.
One thing that people regularly do is quantify how much of a particular activity
they do, but they rarely quantify how well they do it. In this project, your goal
will be to use data from accelerometers on the belt, forearm, arm, and dumbell
of 6 participants. They were asked to perform barbell lifts correctly and
incorrectly in 5 different ways. More information is available from the website
here: http://groupware.les.inf.puc-rio.br/har
(see the section on the Weight Lifting Exercise Dataset). 


Data
==========================================================
The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 


Loading and preprocessing the data.
---------------------------------------------------
The data can be downloaded using the below R script.
```{r download_file,echo=TRUE}
downloadFiles<-function(
    dataURL="", destF="t.csv"
    ){
        if(!file.exists(destF)){
            download.file(dataURL, destF, method="curl")
        }else{
            message("data already downloaded.")
        }
    }
```

```{r activity_data,echo=TRUE}
trainURL<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testURL <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
downloadFiles(trainURL, "pml-training.csv")
downloadFiles(testURL, "pml-test.csv")

training <- read.csv("pml-training.csv")
testing <-read.csv("pml-test.csv")
```

training
```{r}
str(training)
table(training$classe)
```

removing zero covariates
```{r}
library(caret)
nzv <- nearZeroVar(training, saveMetrics=TRUE)
nzv
nzv <- nearZeroVar(training)
training<-training[,-nzv]
```

```{r}
#M <- abs(cor(training[,-c(2,5,100)]))
#diag(M) <- 0
#which(M>0.8, arr.ind=T)
#prComp <- prcomp(training[,-160])
#preProc <- preProcess(training[,-160], method="pca", pcaComp=2)
preProc <- preProcess(training[,-c(2,5,100)], method="pca", thresh=0.8)
trainingPC <- predict(preProc,training[,-c(2,5,100)] )

```

```{r}
modFit <- train(classe~., data=training, method="rpart", prox=TRUE)
modFit$finalModel
```

